Neutron Star Mergers and their Impact on Second Generation Star Formation in the Early Universe

### 2023-09-15 NSM_Skinner_Wise_arxiv/NSM_mixing.tex
Making new directory for arxiv, to keep arxiv version and referee version separate.

% Abstract of the paper
\begin{abstract}
	The exact evolution of elements in the universe, from primordial to heavier elements produced via the r-process, is still under scrutiny. The supernova deaths of the very first stars led to the enrichment of their local environments, and can leave behind neutron stars (NS) as remnants. These remnants can end up in binary systems with other NSs, and eventually merge, allowing for the r-process to occur. We study the scenario where a single NS merger (NSM) enriches a halo early in its evolution to understand the impact on the second generation of stars and their metal abundances. We perform a suite of high resolution cosmological zoom-in simulations using \enzo{} where we have implemented a new NSM model varying the explosion energy and the delay time. In general, a NSM leads to significant r-process enhancement in the second generation of stars \textbf{in a galaxy with a stellar mass of $\sim 10^5 \Ms$ at redshift 10.  A high explosion energy leads to a Pop II mass fraction of 72\% being highly enhanced with r-process elements, while a lower explosion energy leads to 80\% being enhanced, but only 14\% being highly enhanced. When the NSM has a short delay time of 10 Myr, only 5\% of the mass fraction of Pop II stars is highly enhanced, while 64\% is highly enhanced for the longest delay time of 100 Myr.} This work represents a stepping stone towards understanding how NSMs impact their environments and metal abundances of descendant generations of stars.
\end{abstract}

@@

The motivation for this work comes from the r-process enriched stars in Ret II, an ultrafaint dwarf (UFD) galaxy which serves as an ideal laboratory for studying early universe star formation. These systems typically form most of their stars early in the universe, before reionization was able to suppress star formation \citep{Weisz14}. This means that these stars form in the first couple hundred million years after the Big Bang and haven't been affected by recent mergers or enriched by recent star formation. Ret II was detected by the Dark Energy Survey \citep{Bechtol15}, whose stars have since been extensively studied \citep{Simon15, Roederer16, Ji16, Ji22}. Recently, \citet{Ji22} have found that $\sim 72\%$ of the stars in Ret II are r-process enriched, whose source could have been a prompt NSM. Since this galaxy is so isolated, it serves as an ideal case study for simulating this scenario: namely, what happens if a single NSM, produced from the remnants of Pop III stars, occurs within a galaxy in the early universe? \textbf{This type of scenario is particularly unique due to the chemical impact a NSM could have on the star formation history of a galaxy. If a NSM did occur early in the history of Ret II, the r-process enriched stars are a special and direct connection to previous generations of stars.}

@@

\subsection{Neutron star merger model and simulation suite} \label{sec:NSM_model}
\textbf{To explore how a NSM descendent from Pop III stars affects the r-process enrichment of Pop II stars, w}e implement a new NSM model within \enzo{} to control the delay time and the explosion energy of the event, modelled as a single star particle. Upon completion of the original simulation, a single Pop III star is chosen as the candidate NSM event. The chosen Pop III star must be between $20 \leq M / M_{\odot} \leq 30$ so that when this system is split into a symmetric binary system, the initial masses of the component Pop III stars is between $10 \leq M / M_{\odot} \leq 15$, corresponding to the SN models from \citet{Heger10} which will produce a NS as a remnant. We use their model C as a point of reference since the explosion energy, $0.9 \times 10^{51}$ erg, is the closest to the CCSN explosion energy in \enzo{}. The simulation is restarted at the output where the Pop III star forms, indicating which Pop III star and what NSM parameters should be included. We chose the first allowed Pop III star that forms in the simulation to become the source of the NSM event in order to allow for the maximum amount of time between the NSM event and for the second generation of stars to form. The Pop III star that we chose has an initial mass of $28.7 M_{\odot}$ that was randomly selected from an IMF described in \citet{Skinner20}. The chosen Pop III star is then converted into a symmetric Pop III binary system with component masses of $\sim 14.35 M_{\odot}$, but is still modelled as a single star particle with an ionizing hydrogen photon rate of $2.9 \times 10^{48} \mathrm{s}^{-1}$, which is equal to the sum of the two individual stellar photon rates from \citet{Schaerer02}.

@@

A second explosion goes off with $E_{\textrm{NSM}}$ after a time $t_{\textrm{merge}}$, with the parameters chosen in a particular run given the choices in Table \ref{tab:NSM_params}. The final remnant particle is then converted to an inert particle with a mass equal to the sum of the remnant masses. We do not consider any EM radiation from the NSs nor any kilonova effects of a NSM for simplicity; we only follow the mass ejected from this event. We track the r-process material in a passive scalar field which does not affect metal cooling rates. For this proof of concept, the total ejecta mass is set to $0.01 \Ms$, but in practice, due to the Cartesian discretization of a sphere in of Enzo, the actual mass ejected is $0.0169 M_{\odot}$. \textbf{The delay times chosen for these runs are on the shorter end of timescales for the merging of two NSs to take place, but for this proof of concept, we explore this particular situation where a NSM occurs early in the star formation history of the galaxy. For a galaxy like Ret II, a NSM would have had to occur within 500 Myr to allow for ample time for thorough mixing to take place within the halo \citep{Simon23}. Delay times of 10, 100, and 300 Myr are short, but are not completely out of the realm of possibilities, and allow us to see how a NSM affects subsequent star formation at this early point in time.}

@@

A Pop II star forms using the same criteria, except without the molecular hydrogen fraction requirement, and with a metallicity greater than the above value. A Pop II particle represents a cluster of Pop II stars with a minimum cluster mass of 1000 \Ms. There are instances when a Pop II star will form below the minimum mass, in which case the star accretes mass without producing feedback until it reaches that minimum mass. Both Pop III and Pop II stars are assigned a metallicity based on the metallicity fraction of the cell that the particle forms in. Because Pop II stars form as star clusters, the exact number of Pop II star particles that form is not a good indicator of how productive an environment is at forming Pop II stars. Rather, the total Pop II stellar mass is the indicator that will be presented throughout this work. \textbf{We do not take into account the effects of Pop III or Pop II binary systems due to the limiting resolution of our simulations.}

@@

\subsection{Evolution of the main halo \textbf{in the original run}} \label{sec:main_halo}
\begin{figure} 
	\includegraphics[width=\columnwidth]{plots/run_OG_stellar_dist.pdf}
	\caption[Radial distribution of Pop II and III stars in the main halo.]{The radial distribution of Pop III and Pop II stars in the halo in the original run at $z = 10$. The orange and blue lines show the Pop III and Pop II stars, respectively. The black line indicates the virial radius of the halo. The half-mass radius for this halo is 331.2 pc}
	\label{fig:og_stellarmass}
\end{figure}

@@

\textbf{In the original run with no NSM,} the most massive progenitor of the main halo is first identified at $z = 26.9$ with a mass of $10^{4.6} \Ms$, and ends with a mass of $10^{8.19} \Ms$ at $z = 10$. During that time, the halo forms two Pop III stars with masses of 26.5 \Ms{} and 195.5 \Ms{} at 150.7 Myr ($z = 22.6$) and 286.1 Myr ($z = 14.4$), respectively. The halo experiences 25 Pop III SN from stars that accrete into the halo and from the two that form within the main progenitor of the halo. 26 Pop II star particles form within the halo with a total stellar mass of $10^{4.5} \Ms$, and with a median and mean mass weighted creation time of 441 Myr ($z = 10.5$) and 402 Myr ($z = 11.3$), respectively. There are also five Pop II star particles that fall inside the halo. Figure \ref{fig:og_stellarmass} shows the radial distribution of Pop III and Pop II stars within the halo at $z = 10$. The half-mass radius is 331.2 pc, which is on the higher end of results from other simulations of UFD galaxies \citep{Ricotti16}. The Pop II stars all tend to cluster within 1 kpc of the center of the halo, while the Pop III stars are more dispersed throughout the halo.

@@

\begin{figure} 
	\includegraphics[width=\columnwidth]{plots/eng_evolution.pdf}
	\caption[The halo mass, Pop II stellar mass, gas fraction, and r-process mass as a function of time for the explosion energy variation runs.]{The evolution of the main halo in each explosion energy variation run and the original \textbf{(OG)} run. The top panel shows the halo mass, second from the top shows the Pop II stellar mass, third from the top shows the average gas fraction, and the bottom shows the r-process mass. The NSM occurs at 286 Myr.}
	\label{fig:eng_evol}
\end{figure}

@@

\textbf{}

@@

\textbf{For all runs, we see similar Pop III stellar masses as compared to some recent studies. \citet{Riaz22} studied the Pop III contribution to high redshift galaxies for a range of halo masses. They find that a halo with a mass of $10^7$ \Ms{} has a Pop III mass fraction of $\sim 0.1$ at $z \approx 12$. For our runs, at a similar time and halo mass, we also see a Pop III mass fraction of $\sim 0.1$. By $z = 10$, we see $\sim 10^{4}$ \Ms{} of Pop II stars and $\sim 10^{3}$ \Ms{} for Pop III stars, as compared to $10^3 - 10^5$ \Ms{} of Pop II stars and  $10^1 - 10^2$ \Ms{} of Pop III stars found by \citet{Riaz22}. While we do see a slightly higher value of Pop III stellar masses as compared to them, our overall mass fractions are similar. \citet{Yajima23} studied the first galaxies and their Pop III contributions in order to compare with results from JWST. They also look at the Pop III mass fraction (see their Figure 5) as compared to the total stellar mass. They report a much broader range of values for different redshifts in general. For a stellar mass of $10^{4.5}$ \Ms, they find a Pop III mass fraction of $\sim 0.18$, while we see a mass fraction for that stellar mass around 0.03. However, we do see a similar trend, that as stellar mass increases, the Pop III mass fraction begins to decrease. At the end of our simulations when our halo has a stellar mass of $\sim 10^{5.5}$ \Ms, we find a Pop III mass fraction of $\sim 0.006$, which is similar to \citet{Yajima23}. Based on these results, we agree with their conclusion that for halos with stellar masses of $\sim 10^{7}$ \Ms, the Pop III mass fraction is small, at $< 0.01$ }

This work represents some of the first simulations on the chemical and dynamical impacts of a NSM, with various parameters, in a cosmological setting, and thus there are some difficulties that come with this kind of study. The first is that the small sample size prevents us from saying anything conclusive about the NSM parameters in general. While we see some trends across the data, the inherent randomness across simulations leads to different amounts of star formation at different times and locations. On top of that, because we randomly sample the Pop III IMF, if two simulations are looking very similar and both go on to form a Pop III star in a similar location, one can be sampled at a low mass and the other at a very high mass, and affect the surrounding region in completely different ways. What may seem like a small difference in the Pop III stellar mass can go on to have impactful consequences on future star formation. This means that it can be difficult to pick apart why certain stars form where and at what times. We stress that this is merely a proof-of-concept study to show that we can study the impacts of NSMs in a cosmological setting to present meaningful results that can enhance our understanding of r-process enrichment in the early universe. \textbf{We plan to follow up this paper with a study that separates the source of differences between the simulations, and are outside the scope of this paper.}

>>>>

% Abstract of the paper
\begin{abstract}
	The exact evolution of elements in the universe, from primordial to heavier elements produced via the r-process, is still under scrutiny. The supernova deaths of the very first stars led to the enrichment of their local environments, and can leave behind neutron stars (NS) as remnants. These remnants can end up in binary systems with other NSs, and eventually merge, allowing for the r-process to occur. We study the scenario where a single NS merger (NSM) enriches a halo early in its evolution to understand the impact on the second generation of stars and their metal abundances. We perform a suite of high resolution cosmological zoom-in simulations using \enzo{} where we have implemented a new NSM model varying the explosion energy and the delay time. In general, a NSM leads to significant r-process enhancement in the second generation of stars in a galaxy with a stellar mass of $\sim 10^5 \Ms$ at redshift 10.  A high explosion energy leads to a Pop II mass fraction of 72\% being highly enhanced with r-process elements, while a lower explosion energy leads to 80\% being enhanced, but only 14\% being highly enhanced. When the NSM has a short delay time of 10 Myr, only 5\% of the mass fraction of Pop II stars is highly enhanced, while 64\% is highly enhanced for the longest delay time of 100 Myr. This work represents a stepping stone towards understanding how NSMs impact their environments and metal abundances of descendant generations of stars.
\end{abstract}

@@

The motivation for this work comes from the r-process enriched stars in Ret II, an ultrafaint dwarf (UFD) galaxy which serves as an ideal laboratory for studying early universe star formation. These systems typically form most of their stars early in the universe, before reionization was able to suppress star formation \citep{Weisz14}. This means that these stars form in the first couple hundred million years after the Big Bang and haven't been affected by recent mergers or enriched by recent star formation. Ret II was detected by the Dark Energy Survey \citep{Bechtol15}, whose stars have since been extensively studied \citep{Simon15, Roederer16, Ji16, Ji22}. Recently, \citet{Ji22} have found that $\sim 72\%$ of the stars in Ret II are r-process enriched, whose source could have been a prompt NSM. Since this galaxy is so isolated, it serves as an ideal case study for simulating this scenario: namely, what happens if a single NSM, produced from the remnants of Pop III stars, occurs within a galaxy in the early universe? This type of scenario is particularly unique due to the chemical impact a NSM could have on the star formation history of a galaxy. If a NSM did occur early in the history of Ret II, the r-process enriched stars are a special and direct connection to previous generations of stars.

@@

\subsection{Neutron star merger model and simulation suite} \label{sec:NSM_model}
To explore how a NSM descendent from Pop III stars affects the r-process enrichment of Pop II stars, we implement a new NSM model within \enzo{} to control the delay time and the explosion energy of the event, modelled as a single star particle. Upon completion of the original simulation, a single Pop III star is chosen as the candidate NSM event. The chosen Pop III star must be between $20 \leq M / M_{\odot} \leq 30$ so that when this system is split into a symmetric binary system, the initial masses of the component Pop III stars is between $10 \leq M / M_{\odot} \leq 15$, corresponding to the SN models from \citet{Heger10} which will produce a NS as a remnant. We use their model C as a point of reference since the explosion energy, $0.9 \times 10^{51}$ erg, is the closest to the CCSN explosion energy in \enzo{}. The simulation is restarted at the output where the Pop III star forms, indicating which Pop III star and what NSM parameters should be included. We chose the first allowed Pop III star that forms in the simulation to become the source of the NSM event in order to allow for the maximum amount of time between the NSM event and for the second generation of stars to form. The Pop III star that we chose has an initial mass of $28.7 M_{\odot}$ that was randomly selected from an IMF described in \citet{Skinner20}. The chosen Pop III star is then converted into a symmetric Pop III binary system with component masses of $\sim 14.35 M_{\odot}$, but is still modelled as a single star particle with an ionizing hydrogen photon rate of $2.9 \times 10^{48} \mathrm{s}^{-1}$, which is equal to the sum of the two individual stellar photon rates from \citet{Schaerer02}.

@@

A second explosion goes off with $E_{\textrm{NSM}}$ after a time $t_{\textrm{merge}}$, with the parameters chosen in a particular run given the choices in Table \ref{tab:NSM_params}. The final remnant particle is then converted to an inert particle with a mass equal to the sum of the remnant masses. We do not consider any EM radiation from the NSs nor any kilonova effects of a NSM for simplicity; we only follow the mass ejected from this event. We track the r-process material in a passive scalar field which does not affect metal cooling rates. For this proof of concept, the total ejecta mass is set to $0.01 \Ms$, but in practice, due to the Cartesian discretization of a sphere in of Enzo, the actual mass ejected is $0.0169 M_{\odot}$. The delay times chosen for these runs are on the shorter end of timescales for the merging of two NSs to take place, but for this proof of concept, we explore this particular situation where a NSM occurs early in the star formation history of the galaxy. For a galaxy like Ret II, a NSM would have had to occur within 500 Myr to allow for ample time for thorough mixing to take place within the halo \citep{Simon23}. Delay times of 10, 100, and 300 Myr are short, but are not completely out of the realm of possibilities, and allow us to see how a NSM affects subsequent star formation at this early point in time.

@@

A Pop II star forms using the same criteria, except without the molecular hydrogen fraction requirement, and with a metallicity greater than the above value. A Pop II particle represents a cluster of Pop II stars with a minimum cluster mass of 1000 \Ms. There are instances when a Pop II star will form below the minimum mass, in which case the star accretes mass without producing feedback until it reaches that minimum mass. Both Pop III and Pop II stars are assigned a metallicity based on the metallicity fraction of the cell that the particle forms in. Because Pop II stars form as star clusters, the exact number of Pop II star particles that form is not a good indicator of how productive an environment is at forming Pop II stars. Rather, the total Pop II stellar mass is the indicator that will be presented throughout this work. We do not take into account the effects of Pop III or Pop II binary systems due to the limiting resolution of our simulations.

@@

\subsection{Evolution of the main halo in the original run} \label{sec:main_halo}
\begin{figure} 
	\includegraphics[width=\columnwidth]{plots/run_OG_stellar_dist.pdf}
	\caption[Radial distribution of Pop II and III stars in the main halo.]{The radial distribution of Pop III and Pop II stars in the halo in the original run at $z = 10$. The orange and blue lines show the Pop III and Pop II stars, respectively. The black line indicates the virial radius of the halo. The half-mass radius for this halo is 331.2 pc}
	\label{fig:og_stellarmass}
\end{figure}

@@

In the original run with no NSM, the most massive progenitor of the main halo is first identified at $z = 26.9$ with a mass of $10^{4.6} \Ms$, and ends with a mass of $10^{8.19} \Ms$ at $z = 10$. During that time, the halo forms two Pop III stars with masses of 26.5 \Ms{} and 195.5 \Ms{} at 150.7 Myr ($z = 22.6$) and 286.1 Myr ($z = 14.4$), respectively. The halo experiences 25 Pop III SN from stars that accrete into the halo and from the two that form within the main progenitor of the halo. 26 Pop II star particles form within the halo with a total stellar mass of $10^{4.5} \Ms$, and with a median and mean mass weighted creation time of 441 Myr ($z = 10.5$) and 402 Myr ($z = 11.3$), respectively. There are also five Pop II star particles that fall inside the halo. Figure \ref{fig:og_stellarmass} shows the radial distribution of Pop III and Pop II stars within the halo at $z = 10$. The half-mass radius is 331.2 pc, which is on the higher end of results from other simulations of UFD galaxies \citep{Ricotti16}. The Pop II stars all tend to cluster within 1 kpc of the center of the halo, while the Pop III stars are more dispersed throughout the halo.

@@

\begin{figure} 
	\includegraphics[width=\columnwidth]{plots/eng_evolution.pdf}
	\caption[The halo mass, Pop II stellar mass, gas fraction, and r-process mass as a function of time for the explosion energy variation runs.]{The evolution of the main halo in each explosion energy variation run and the original (OG) run. The top panel shows the halo mass, second from the top shows the Pop II stellar mass, third from the top shows the average gas fraction, and the bottom shows the r-process mass. The NSM occurs at 286 Myr.}
	\label{fig:eng_evol}
\end{figure}

@@

@@

For all runs, we see similar Pop III stellar masses as compared to some recent studies. \citet{Riaz22} studied the Pop III contribution to high redshift galaxies for a range of halo masses. They find that a halo with a mass of $10^7$ \Ms{} has a Pop III mass fraction of $\sim 0.1$ at $z \approx 12$. For our runs, at a similar time and halo mass, we also see a Pop III mass fraction of $\sim 0.1$. By $z = 10$, we see $\sim 10^{4}$ \Ms{} of Pop II stars and $\sim 10^{3}$ \Ms{} for Pop III stars, as compared to $10^3 - 10^5$ \Ms{} of Pop II stars and  $10^1 - 10^2$ \Ms{} of Pop III stars found by \citet{Riaz22}. While we do see a slightly higher value of Pop III stellar masses as compared to them, our overall mass fractions are similar. \citet{Yajima23} studied the first galaxies and their Pop III contributions in order to compare with results from JWST. They also look at the Pop III mass fraction (see their Figure 5) as compared to the total stellar mass. They report a much broader range of values for different redshifts in general. For a stellar mass of $10^{4.5}$ \Ms, they find a Pop III mass fraction of $\sim 0.18$, while we see a mass fraction for that stellar mass around 0.03. However, we do see a similar trend, that as stellar mass increases, the Pop III mass fraction begins to decrease. At the end of our simulations when our halo has a stellar mass of $\sim 10^{5.5}$ \Ms, we find a Pop III mass fraction of $\sim 0.006$, which is similar to \citet{Yajima23}. Based on these results, we agree with their conclusion that for halos with stellar masses of $\sim 10^{7}$ \Ms, the Pop III mass fraction is small, at $< 0.01$.

This work represents some of the first simulations on the chemical and dynamical impacts of a NSM, with various parameters, in a cosmological setting, and thus there are some difficulties that come with this kind of study. The first is that the small sample size prevents us from saying anything conclusive about the NSM parameters in general. While we see some trends across the data, the inherent randomness across simulations leads to different amounts of star formation at different times and locations. On top of that, because we randomly sample the Pop III IMF, if two simulations are looking very similar and both go on to form a Pop III star in a similar location, one can be sampled at a low mass and the other at a very high mass, and affect the surrounding region in completely different ways. What may seem like a small difference in the Pop III stellar mass can go on to have impactful consequences on future star formation. This means that it can be difficult to pick apart why certain stars form where and at what times. We stress that this is merely a proof-of-concept study to show that we can study the impacts of NSMs in a cosmological setting to present meaningful results that can enhance our understanding of r-process enrichment in the early universe. We plan to follow up this paper with a study that separates the source of differences between the simulations, and are outside the scope of this paper.
